Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-06-10 paper/paper.tex
more fixups

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings\suggestremove{ in animals and humans}.
\end{abstract}

@@

The goal of the RL agent is then to update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be characteristic for a certain agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning of long-term goal that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays for higher rewards.
Such an agent puts relatively more emphasis on rewards expected in
a longer-term future.
Precisely, rewards that are not expected to come within\footnote{\textbf{N.B.:} $\frac{1}{1-\gamma} = 1 + \gamma + \gamma^2 + \gamma^3 + \ldots$}
$\tau := 1/(1 - \gamma)$
time steps from the present point are disregarded. This reduces the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the farsighted future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

\begin{equation}
  \pi^{\text{hard-max}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
  The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
  $Q$-values. Such a mapping from states to actions is conventionally called \textit{policy matrix}\citep{mnih2015,silver2016mastering}.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries. \suggestadd{An additional layer of learning concerns the addition of new entries in the table (i.e. state and action). The framing of the policy matrix is not described here but could be supported by synaptic epigenesis \citep{gisiger_acquisition_2005}. Indeed, the tuning of synaptic weights through learning can stabilize additional patterns of activity by
creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors will then constrain both
the number of factors taken into account by decision processes and the number of behaviors at reach by the agent \citep{wang_decision_2008}.}

@@

\suggestadd{The principal connectivity gradient in the cortex appears to be greatly expanded in humans compare to macaques, suggesting a phylogenetically conserved axis of cortical expansion with the DMN emerging at the extreme-end in humans \citep{margulies_situating_2016}. Neurocomputational models of dyadic whole brain demonstrated how the human connectivity topology, on top of facilitating processing at the intra-individual level, can explain our propensity to coordinate through sensorimotor loops with others at the inter-individual level \citep{dumas_anatomical_2012}. DMN is moreover largely overlapping with neural networks associated with higher level social cognition \citep{schilbach_introspective_2012}. For instance, the ventromedial prefrontal cortex, the posterior cingulate/precuneus, and the temporo-parietal junction (rTPJ) play together a key role in bridging the gap between self and other by
integrating low-level embodied processes within higher level inference-based mentalizing \citep{lombardo_shared_2009}.}

@@

\subsection{The free-energy principle and active inference}
% INTRO
According to the free-energy principle (FEP) and
active inference theories \citep{friston2010free, fristonAIorRL},
the brain corresponds to
a biomechanical inference engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --or more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from the brain's internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
Precursors of this theory can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics. The model then allows
to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
The FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$ to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
while the generative model tries to produce sensations $\x$ which match these
imagined mental scenarios.
Howewer, surprise is challenging to optimize numerically
because we need to sum over all hidden causes of the sensations.
Instead, FEP therefore minimizes an upper-bound of surprise,
namely the \text{free-energy} $F_G^R(\x)$ given by

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which we outline in the following.
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As previously noted \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it postulates a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
Thus, the brain is unlikely to implement a variant of the wake-sleep algorithm.
We note that the recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. This is because VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free-energy, VAEs can be efficiently trained via back-propagation of prediction errors.

Since FEP-based theories \citep{friston2010free,fristonAIorRL} conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
in this environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a closed space is used as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddles to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would readily
identify the visual-input-free room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply leaving the room.

>>>>

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

The goal of the RL agent is then to update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be characteristic for a certain agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning of long-term goal that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays for higher rewards.
Such an agent puts relatively more emphasis on rewards expected in
a longer-term future.
Precisely, rewards that are not expected to come within
% \footnote{\textbf{N.B.:} $\frac{1}{1-\gamma} = 1 + \gamma + \gamma^2 + \gamma^3 + \ldots$}
$\tau := 1/(1 - \gamma)$
time steps from the present point are disregarded. This reduces the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the farsighted future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

\begin{equation}
  \pi^{\text{hard-max}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
parameters $\theta$ would correspond to synaptic weights and connection strengths between
brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
$Q$-values. Such a mapping from states to actions is conventionally called \textit{policy matrix}\citep{mnih2015,silver2016mastering}.
Learning consists in starting from a given table and
updating it during action choices,
which take the agent to different table entries.
An additional layer of learning concerns the addition of new entries in the table (i.e., state and action). The framing of the policy matrix is not described here but could be supported by synaptic epigenesis \citep{gisiger_acquisition_2005}. Indeed, the tuning of synaptic weights through learning can stabilize additional patterns of activity by
creating new attractors in the neural dynamics landscape \citep{takeuchi_synaptic_2014}. Those attractors will then constrain both
the number of factors taken into account by decision processes and the number of behaviors at reach by the agent \citep{wang_decision_2008}.}

@@

The principal connectivity gradient in the cortex appears to be greatly expanded in humans compare to macaques, suggesting a phylogenetically conserved axis of cortical expansion with the DMN emerging at the extreme-end in humans \citep{margulies_situating_2016}. Neurocomputational models of dyadic whole brain demonstrated how the human connectivity topology, on top of facilitating processing at the intra-individual level, can explain our propensity to coordinate through sensorimotor loops with others at the inter-individual level \citep{dumas_anatomical_2012}. DMN is moreover largely overlapping with neural networks associated with higher level social cognition \citep{schilbach_introspective_2012}. For instance, the ventromedial prefrontal cortex, the posterior cingulate/precuneus, and the temporo-parietal junction (rTPJ) play together a key role in bridging the gap between self and other by
integrating low-level embodied processes within higher level inference-based mentalizing \citep{lombardo_shared_2009}.

@@

\subsection{The free-energy principle and active inference}
% INTRO
According to the free-energy principle (FEP) and
active inference theories \citep{friston2010free, fristonAIorRL},
the brain corresponds to
a biomechanical inference engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --or more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from the brain's internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
Precursors of this theory can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ of objects in the world, their interactions and dynamics. The model then allows
to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
The FEP also postulates a companion model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$ to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
while the generative model tries to produce sensations $\x$ which match these
imagined mental scenarios.
Howewer, surprise is challenging to optimize numerically
because we need to sum over all hidden causes of the sensations.
Instead, FEP therefore minimizes an upper-bound of surprise,
namely the \text{free-energy} $F_G^R(\x)$ given by

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which we outline in the following.
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As previously noted \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it postulates a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
Thus, the brain may therefore be unlikely to implement a variant of the wake-sleep algorithm.
We note that the recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. This is because VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free-energy, VAEs can be efficiently trained via back-propagation of prediction errors.

Since FEP-based theories \citep{friston2010free,fristonAIorRL} conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
in this environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a closed space is used as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddles to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would readily
identify the visual-input-free room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.
